Optimal updating factor in adaptive flat-distribution simulations

### 2016-04-22 doc/invt.tex
invt.tex move sinc.gp

@@

In the above, we have focused on
the asymptotic error.
%
For a finite-length simulation,
we should include the residual part
and minimize the total error.
%
For the bandpass updating schemes,
this leads to the optimal value
of the free parameter, $t_0$,
which is, according to Eq. \eqref{eq:half_alpha0},
%
%%Assuming the system is initially equilibrated
%%under a constant updating magnitude,
%%$\alpha_0$,
%Assuming Eq. \eqref{eq:error_res1} with
%$$
%q(T)
%=
%\int_0^T \frac{ dt } { t + t_0 }
%=
%\ln \frac{ T + t_0 } { t_0 }.
%$$
%%
%we have, from Eqs. \eqref{eq:error_split}
%and \eqref{eq:error_asym_sinc},
%%
%\begin{equation}
%  E
%  =
%  E_R + E_A
%  =
%  \frac { \frac 1 2 \alpha_0 \, t_0^2 + T }
%        {          ( T + t_0 )^2          }
%  \sum_{ k = 2 }^K \Gamma_k
%  .
%\notag
%%\label{eq:error_sinc0}
%\end{equation}
%%
%The minimum is reached at
%
\begin{equation}
  t_0
  =
  \frac{     2    }
       { \alpha_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
and at this point
%
\begin{equation}
  E
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 2 }^K
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
In Appendix \ref{sec:equilerr},
we show that the above error tends
to the error of the ideal equilibrium FDS simulation
in the asymptotic limit.

@@

Our test system is
a model one-dimensional system
of $n = 100$ bins,
with a flat target distribution
$p_i = 1/n$.
%
%Since in the asymptotic regime,
%the distribution converges to $p_i$
%no matter the intrinsic distribution, $p^*_i$,
%we also set the latter to be flat for convenience.
The intrinsic distribution, $p^*_i$,
is also assumed to be flat.
%
Unless specified otherwise,
we assume $i$ to be non-periodic.
%
We wish to see how
the schedule, $\alpha(t)$,
affects the fluctuating error
of the bias potential
at the end of the simulation.

Initially,
we set the bias potential to zero,
$V_i \equiv 0$,
%
and equilibrate the system for $10^7$ steps
under a constant updating magnitude,
$\alpha_0 = 10^{-4}$,
unless specified otherwise.
%
After this,
we reset the origin of the time, $t$,
and started the testing schedule.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

Since the error depends on
the underlying sampling process,
we tested two sampling processes,
a global one and a local one.
%
Both are based on
Metropolis MC algorithm\cite{
  metropolis1953, newman, frenkel,
  landau_binder}:
%
in each step, a new bin index $j$ is proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(V_i - V_j) \}.
$
However,
in the first or global sampling process,
we choose the destination $j$
uniformly out of the $n$ bins.
%
This process
emulates the perfect sampling process
in Sec. \ref{sec:Gamma} asymptotically.
%
In the second or local sampling process,
$j$ is either $i - 1$ or $i + 1$
(adjusted by the boundary condition)
with equal probability $1/2$,
which mimics the one-step sampling process
in Sec. \ref{sec:Gamma}.
%

@@

We now turn to the optimal schedule
predicted by Eq. \eqref{eq:q_opt}.
%
For the Gaussian updating scheme
with $\sigma = 10$,
we computed the schedule for the local sampling process,
and compared it with the inverse-time schedule
Eq. \eqref{eq:alpha_invtlambda}
with the optimal $\lambda$.
%
As shown in Fig. \ref{fig:optacmp}(a),
the optimal schedule is a more complex curve
than the optimized inverse-time schedule.
%
The complexity of the optimal schedule
corresponds to the non-exponential tail
of the mass distribution function,
as shown Fig. \ref{fig:optacmp}(b)
[cf. Sec. \ref{sec:mass_distr}].
%
The errors in the two cases were
$9.1 \times 10^{-5}$
and
$1.7 \times 10^{-4}$,
respectively,
%for the local sampling process,
showing the superiority of
the optimal schedule from Eq. \eqref{eq:q_opt}.
%
Besides, we verified the scaling property
discussed in Sec. \ref{sec:mass_distr},
by comparison with a simulation [Fig. \ref{fig:optacmp}(a)]
of half length, $T' = 5 \times 10^7$
and doubled initial updating magnitude,
$\alpha_0 = 2 \times 10^{-4}$,
which left the normalized mass distribution invariant,
as shown in Fig. \ref{fig:optacmp}(b).

@@

On the other hand,
the optimal schedules of the bandpass updating schemes
are always given by the inverse-time formula,
Eq. \eqref{eq:alpha_invt1},
with $t_0$ given by Eq. \eqref{eq:t0_sinc},
as shown in Fig. \ref{fig:sinc}(a).
%
In Fig. \ref{fig:sinc}(b),
we show that the updating kernels
of the bandpass updating schemes
are similar to Gaussian ones
except the oscillatory tails.
%
The equivalent width, $\sigma$, is roughly
%
\begin{equation}
  \sigma
  \approx
  \frac
  {
    n
  }
  {
    \sqrt{ 2 \, \pi } \, g \, K
  }
  ,
\label{eq:sigma_equiv}
\end{equation}
%
where $g = 2$ for the periodic case
or $g = 1$ otherwise.
%
The kernel is a zigzag in the non-periodic case
because of the term $(-1)^{n+K+l-1}$
in Eq. \eqref{eq:mu_sinc_refl}.
%
However, as shown in the inset, % of Fig. \ref{fig:sinc}(b),
the ruggedness disappears
in the updating matrix by Eq. \eqref{eq:w_band_refl},
as the above oscillatory term for $l = i - j$
is canceled by the term for either $l = i + j - 1$
or $l = i + j - 2 \, n - 1$
with the opposite sign.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/sinc.pdf}
  }
  \caption{
    \label{fig:sinc}
    Bandpass updating schemes.
    (a) Optimal schedule for a periodic variable
    with $K = 3$,
    and that for a non-periodic variable
    with $K = 6$.
    (b) Updating kernels
    compared to a Gaussian with a similar width,
    $\sigma = 6$.
    %
    The local sampling process was used.
    The lines are a guide to the eyes.
    Inset. Updating matrix in the non-periodic case
    is free of the zigzaggedness shown in the kernel.
    %
    \note{This figure was produced by \texttt{doc/fig/sinc.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h} for panel (a),
      and the functions \texttt{savewin()}
      and \texttt{savewinmat()}
      in \texttt{prog/invt.h} for panel (b),
      The data were saved under the directory \texttt{data/sinc}.
      See the \texttt{make} object \texttt{sincsingle}
      in the \texttt{Makefile} there.
    }%
  }
\end{center}
\end{figure}

@@

%To use Eq. \eqref{eq:q_opt} in a real simulation,
%we need to know both $\lambda_k$'s and $\Gamma_k$'s.
%
We can also
use Eq. \eqref{eq:y2_eql}.
measure $\Gamma_k$'s in
a simulation with a constant schedule,
$\alpha(t) = \alpha_0$.
%
%A practical difficulty is that
However,
we often do not know the intrinsic distribution, $p^*_i$,
in advance to compute
the shifted bias potential, $v_i$,
from the original one, $V_i$
(cf. Sec. \ref{sec:FDS}).
%
To circumvent the problem, we
modify Eq. \eqref{eq:y2_eql} as
%
\begin{equation}
  \operatorname{var} Y_k
  =
  \frac{1}{2}
  \alpha \, \Gamma_k \, \lambda_k,
\label{eq:varY}
\end{equation}
%
where
``$\operatorname{var}$''
means the variance from time averaging,
and
%
\begin{align}
Y_k
&=
\sum_{ i = 1 }^n
  \phi_{k i} \, (V_i - \bar V),
\label{eq:Y_def}
\end{align}
%
with
$
\bar V
\equiv
\sum_{ j = 1 }^n p_j \, V_j.
$

@@

As an example, we used the method to compute
the integrals, $\Gamma_k$'s, from an MD simulation,
and compared them to those from
the local MC sampling process
[cf. Sec. \ref{sec:results_system}].
%
As shown in Fig. \ref{fig:gamcmp},
the larger values of $\Gamma_k$'s
with smaller $k$'s
from the two sampling processes
were almost proportional,
suggesting similarity
of local sampling processes.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/gamcmp.pdf}
  }
  \caption{
    \label{fig:gamcmp}
    Integrals of the autocorrelation functions
    from an MD simulation versus those
    from the local MC simulation.
    %
    The MD simulation was ran for $10^{10}$ steps with
    the step size $\Delta t_\mathrm{MD} = 0.005$.
    %
    The bin size $\Delta z = 2\, \pi /n$ with $n = 100$,
    and the temperature was $1.0$.
    %
    The velocity rescaling\cite{bussi2007} thermostat was
    applied every MD step with viscocity 20.
    %
    The reflective boundary condition was used.
    %
    \note{The figure was produced by \texttt{doc/fig/gamcmp.gp}.
      Thee data were saved under \texttt{data/gamcmp}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

Here we give a program of computing
the optimal schedule and the error
of a given adaptive FDS simulation.
%
Basically, we need to extract
$\lambda_i$'s and $\Gamma_i$'s
from the input,
and then feed them to Eq. \eqref{eq:q_opt}
for the schedule,
as shown in Fig. \ref{fig:vardep}.
%
%The dependence of variables are shown in Fig. \ref{fig:vardep}.
%
While the updating scheme alone
determines $\lambda_k$'s,
$\Gamma_k$'s depend also on the system
and the underlying sampling process (MC or MD),
and need to be measured using the technique
in Sec. \ref{sec:Gamma_measure}.
%
The complete program is the following.

@@

\item
Run a preliminary adaptive FDS simulation
using the updating scheme with a constant
updating magnitude, $\alpha_0$.

>>>>

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/sinc.pdf}
  }
  \caption{
    \label{fig:sinc}
    Updating kernels of the bandpass updating schemes
    compared to a Gaussian with a similar width.
    %
    The lines are a guide to the eyes.
    Inset. Updating matrix in the non-periodic case.
    %
    \note{This figure was produced by \texttt{doc/fig/sinc.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h} for panel (a),
      and the functions \texttt{savewin()}
      and \texttt{savewinmat()}
      in \texttt{prog/invt.h} for panel (b),
      The data were saved under the directory \texttt{data/sinc}.
      See the \texttt{make} object \texttt{sincsingle}
      in the \texttt{Makefile} there.
    }%
  }
\end{center}
\end{figure}

In Fig. \ref{fig:sinc},
we give examples of the updating kernels
of the bandpass updating schemes,
compared to a Gaussian
with a similar width, $\sigma$,
%
\begin{equation}
  \sigma
  \approx
  \frac
  {
    n
  }
  {
    \sqrt{ 2 \, \pi } \, g \, K
  }
  .
\label{eq:sigma_equiv}
\end{equation}
%
%where $g = 2$ for the periodic case
%or $g = 1$ otherwise.
%
The kernel in the non-periodic case is a zigzag
because of the term $(-1)^{n+K+l-1}$
in Eq. \eqref{eq:mu_sinc_refl}.
%
However, as shown in the inset, % of Fig. \ref{fig:sinc}(b),
the ruggedness disappears
in the updating matrix by Eq. \eqref{eq:w_band_refl},
as the above oscillatory term for $l = i - j$
is canceled by the term for either $l = i + j - 1$
or $l = i + j - 2 \, n - 1$
with the opposite sign.

@@

In the above, we have focused on
the asymptotic error.
%
By including the residual error
and minimize the total,
we get the optimal $t_0$,
which is, according to Eq. \eqref{eq:half_alpha0},
%
%%Assuming the system is initially equilibrated
%%under a constant updating magnitude,
%%$\alpha_0$,
%Assuming Eq. \eqref{eq:error_res1} with
%$$
%q(T)
%=
%\int_0^T \frac{ dt } { t + t_0 }
%=
%\ln \frac{ T + t_0 } { t_0 }.
%$$
%%
%we have, from Eqs. \eqref{eq:error_split}
%and \eqref{eq:error_asym_sinc},
%%
%\begin{equation}
%  E
%  =
%  E_R + E_A
%  =
%  \frac { \frac 1 2 \alpha_0 \, t_0^2 + T }
%        {          ( T + t_0 )^2          }
%  \sum_{ k = 2 }^K \Gamma_k
%  .
%\notag
%%\label{eq:error_sinc0}
%\end{equation}
%%
%The minimum is reached at
%
\begin{equation}
  t_0
  =
  \frac{     2    }
       { \alpha_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
and at this point
%
\begin{equation}
  E
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 2 }^K
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
In Appendix \ref{sec:equilerr},
we show that the above error tends
to the error of the ideal equilibrium FDS simulation
in the asymptotic limit.

@@

In our test system,
$i$ is the index of a one-dimensional
non-periodic variable,
and $n = 100$.
%
Both the target and intrinsic distributions
are assumed to be flat:
$p^*_i = p_i = 1/n$.
%
%We wish to see how
%the schedule, $\alpha(t)$,
%affects the fluctuating error
%of the bias potential
%at the end of the simulation.
%
Initially,
we set the bias potential to zero,
$V_i \equiv 0$,
%
and equilibrate the system for $10^7$ steps
under a constant updating magnitude,
$\alpha_0 = 10^{-4}$,
unless specified otherwise.
%
After this,
we reset the origin of the time, $t$,
and start the testing schedule.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

%Since the error depends on
%the underlying sampling process,
We tested two sampling processes.
%a global one and a local one.
%
Both are based on
Metropolis MC algorithm\cite{
  metropolis1953, newman, frenkel,
  landau_binder}:
%
in each step, a new bin index $j$ is proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(V_i - V_j) \}.
$
However,
in the first or global sampling process,
we choose the destination $j$
uniformly out of the $n$ bins.
%
In the second or local sampling process,
$j$ is either $i - 1$ or $i + 1$
(adjusted by the boundary condition)
with equal probability $1/2$.
%
Asymptotically,
the above global and local sampling processes
emulate the perfect and one-step processes
discussed in Appendix \ref{sec:Gamma},
respectively,
and we shall use
Eqs. \eqref{eq:Gamma_perfect}
and \eqref{eq:Gamma_onestep}
for the respective $\Gamma_k$'s.
%

@@

We computed the optimal schedule from Eq. \eqref{eq:q_opt}
for the Gaussian (with $\sigma = 10$)
and bandpass (with $K = 4$) updating schemes.
%
As shown in Fig. \ref{fig:optacmp},
while the optimal schedules of the bandpass scheme
were always given by the inverse-time formula,
those of the Gaussian updating scheme was complex
and dependent on the sampling process.

@@

@@

%To use Eq. \eqref{eq:q_opt} in a real simulation,
%we need to know both $\lambda_k$'s and $\Gamma_k$'s.
%
For an unknown system,
we can use Eq. \eqref{eq:y2_eql} to
measure $\Gamma_k$'s in
an adaptive FDS simulation
with a constant schedule,
$\alpha(t) = \alpha_0$.
%
%A practical difficulty is that
However,
without the intrinsic distribution, $p^*_i$,
we cannot compute the shifted bias potential, $v_i$,
from the original one, $V_i$
(cf. Sec. \ref{sec:FDS}).
%
To circumvent the problem, we
modify Eq. \eqref{eq:y2_eql} as
%
\begin{equation}
  \operatorname{var} Y_k
  =
  \frac{1}{2}
  \alpha \, \Gamma_k \, \lambda_k,
\label{eq:varY}
\end{equation}
%
where
``$\operatorname{var}$''
means the variance from time averaging,
and
%
\begin{align}
Y_k
&=
\sum_{ i = 1 }^n
  \phi_{k i} \, (V_i - \bar V),
\label{eq:Y_def}
\end{align}
%
with
$
\bar V
\equiv
\sum_{ j = 1 }^n p_j \, V_j.
$

@@

As an example, we used the method to compute
the integrals, $\Gamma_k$'s, from an MD simulation,
and compared them to those from
Eq. \eqref{eq:Gamma_onestep}.
%the one-step sampling process.
%
As shown in Fig. \ref{fig:gamcmp},
the larger values of $\Gamma_k$'s
with smaller $k$'s
from the two sampling processes
were almost proportional,
suggesting similarity
of local sampling processes.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/gamcmp.pdf}
  }
  \caption{
    \label{fig:gamcmp}
    Integrals of the autocorrelation functions
    from an MD simulation versus those
    from the one-step process.
    %
    The MD simulation was ran for $10^{10}$ steps with
    the step size $\Delta t_\mathrm{MD} = 0.005$.
    %
    The bin size $\Delta z = 2\, \pi /n$ with $n = 100$,
    and the temperature was $1.0$.
    %
    The velocity rescaling\cite{bussi2007} thermostat was
    applied every MD step with viscocity 20.
    %
    The reflective boundary condition was used.
    %
    \note{The figure was produced by \texttt{doc/fig/gamcmp.gp}.
      Thee data were saved under \texttt{data/gamcmp}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

Here we give a program of computing
the optimal schedule and the error
of a given adaptive FDS simulation.
%
Basically, we need to extract
$\lambda_i$'s and $\Gamma_i$'s
from the input,
and then feed them to Eq. \eqref{eq:q_opt}
for the schedule,
as shown in Fig. \ref{fig:vardep}.
%
%The dependence of variables are shown in Fig. \ref{fig:vardep}.
%
While the updating scheme alone
determines $\lambda_k$'s,
$\Gamma_k$'s depend also on the system
and the underlying sampling process (MC or MD).
%
By using the technique in Sec. \ref{sec:Gamma_measure}
to measure $\Gamma_k$'s,
we have the following program.

@@

\item
Run a preliminary adaptive FDS simulation
%using the updating scheme
with a constant updating magnitude, $\alpha_0$.
